---
title: "Is automatic or manual transmission better for MPG?"
author: "Chris Lill"
output: pdf_document
---

# Executive Summary

Regression modelling shows that weight and horsepower have the greatest impact on miles per gallon (MPG). Holding these constant, there is a 95% probability that the effect of the transmission is between -2.8 and 0.9 MPG. There is no significant difference between MPG for an automatic and manual transmission.


# Exploratory Data Analysis
The boxplot below shows that there is a correlation between transmission and miles per gallon. However, this result may be explained by other variables so we need to determine which have the best correlation. 

The scatterplot below shows a good correlation between weight(wt) and mpg. It also shows that weight and transmission are correlated, so automatic gearboxes tend to be used on heavier cars. 

The table in appendix 1 shows that in a linear regression model of mpg containing all other factors, weight is the most significant. There is only a 6% probability that it has no effect on mpg.

```{r Exploratory, echo=FALSE, out.width=".5\\textwidth"}
require(ggplot2, quietly = TRUE)
data(mtcars)

mtcars$am <- factor(mtcars$am, labels = c("automatic", "manual"))
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs, labels = c("v", "straight"))
mtcars$gear <- factor(mtcars$gear)
mtcars$carb <- factor(mtcars$carb)
g1 <- ggplot(mtcars, aes(x = factor(am), y = mpg)) +
  geom_boxplot() +
  ggtitle("Correlation between Transmission and MPG")
g1

g2 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(aes(colour = am)) +
  ggtitle("MPG by Weight and Transmission")
g2

```


# Model Selection
To select the most appropriate model, we will select `wt` as the first factor and add other factors incrementally. We'll build a new linear regression model each time and use Anova to see whether the new factor is a significant improvement.

The table below shows that the models with the most significant additional factors are `mpg ~ wt + hp` and `mpg ~ wt * hp`. We will select the `mpg ~ wt + hp` model so that we avoid overfitting on a dataset that only has 32 observations. 

Figure 3 in the Appendix shows residual diagnostics for this model, which shows that this is a reasonable model given the small dataset, however the fit is not perfect and the residuals are not normally distributed with a couple of outliers.


# Results
## Is an automatic or manual transmission better for MPG
To visualise the difference between an automatic and manual transmission we can plot the residuals from our model against transmission(am). 

Out hypothesis is that there is a difference between the means of MPG for automatic and manual cars. This can be carried out with a two sided t-Test assuming equal variance. Note that the correct calculation would have 28 degrees of freedom, subtracting 2 for the model and 2 for the t-test. However the results are so clear that we don't need to do the additional math.

The results show a 95% confidence interval for the difference between -2.8 and 0.9, with a p-value of 0.32. Therefore we fail to reject the null hypothesis, and conclude that there is no significant difference between automatic and manual transmission.

```{r Result1}
fit.wt.hp <- lm(mpg ~ wt + hp, data = mtcars)

boxplot(resid(fit.wt.hp) ~ mtcars$am)
title("Effect of transmission" )

t <- t.test(resid(fit.wt.hp) ~ mtcars$am, var.equal = TRUE)
t$conf[1:2]
```

## Quantify the MPG difference between automatic and manual transmissions
Assuming our model of `mpg ~ wt + hp`, the difference between the mean of the residuals for the groups is 0.9 mpg, with automatic transmissions showing as marginally better.

```{r Result2}
t$estimate["mean in group automatic"] - t$estimate["mean in group manual"]
```


\pagebreak

# Appendix
## Appendix 1: Regression model for mpg with all factors
```{r AllTerms, echo = FALSE}
require(knitr, quietly = TRUE)

data(mtcars)
fit.all <- lm(mpg ~ ., data = mtcars)
kable(summary(fit.all)$coef[,c(1,4)], 
      digits = 2,
      col.names = c("Coefficient", "P-value"),
      caption = "Coefficients and significance for each factor")
mtcars$am <- factor(mtcars$am, labels = c("automatic", "manual"))
mtcars$cyl <- factor(mtcars$cyl)
```

## Appendix 2: Comparison of different regression models for mpg
```{r ModelSelection, echo = FALSE}
suppressMessages(require(dplyr, quietly = TRUE))

# First, generate the models for comparison
fit.wt <- lm(mpg ~ wt, data = mtcars)
fit.wt.am <- lm(mpg ~ wt + am, data = mtcars)
fit.wt.cyl <- lm(mpg ~ wt + factor(cyl), data = mtcars)
fit.wt.disp <- lm(mpg ~ wt + disp, data = mtcars)
fit.wt.hp <- lm(mpg ~ wt + hp, data = mtcars)
fit.wtxhp <- lm(mpg ~ wt * hp, data = mtcars)
fit.wt.hp.am <- lm(mpg ~ wt + hp + am, data = mtcars)
fit.wt.hp.cyl <- lm(mpg ~ wt + hp + factor(cyl), data = mtcars)
fit.wt.hp.disp <- lm(mpg ~ wt + hp + disp, data = mtcars)
fit.wt.hp.drat <- lm(mpg ~ wt + hp + drat, data = mtcars)
fit.wt.hp.qsec <- lm(mpg ~ wt + hp + qsec, data = mtcars)
fit.wt.hp.vs <- lm(mpg ~ wt + hp + factor(vs), data = mtcars)
fit.wt.hp.gear <- lm(mpg ~ wt + hp + factor(gear), data = mtcars)
fit.wt.hp.carb <- lm(mpg ~ wt + hp + factor(carb), data = mtcars)



alt.terms <-c ("mpg ~ wt + am",
              "mpg ~ wt + factor(cyl)", 
              "mpg ~ wt + disp", 
              "mpg ~ wt + hp", 
              "mpg ~ wt * hp",
              "mpg ~ wt + hp + am",
              "mpg ~ wt + hp + factor(cyl)",
              "mpg ~ wt + hp + disp",
              "mpg ~ wt + hp + drat",
              "mpg ~ wt + hp + qsec",
              "mpg ~ wt + hp + factor(vs)",
              "mpg ~ wt + hp + factor(gear)",
              "mpg ~ wt + hp + factor(carb)")
base.terms <- c(rep("mpg ~ wt", 4), rep("mpg ~ wt + hp", 9))
p.value <- c(anova(fit.wt, fit.wt.am)$`Pr(>F)`[2],
             anova(fit.wt, fit.wt.cyl)$`Pr(>F)`[2],
             anova(fit.wt, fit.wt.disp)$`Pr(>F)`[2],
             anova(fit.wt, fit.wt.hp)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wtxhp)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.am)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.cyl)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.disp)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.drat)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.qsec)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.vs)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.gear)$`Pr(>F)`[2],
             anova(fit.wt.hp, fit.wt.hp.carb)$`Pr(>F)`[2])
model.signif <- data.frame(alt.terms, base.terms, p.value)

kable(model.signif, 
      digits = 3,
      col.names = c("Possible model", "Compared with", "P-value"),
      caption = "Significance of additional factor in each model")
```

## Appendix 3 - Residual diagnostics for mpg ~ wt + hp
```{r ResidualDiagnostics, echo = FALSE, fig.height=8}
par(mfrow = c(2, 2))
plot(fit.wt.hp)
```

